\section*{Recommended Follow-Up Annotation Questions for Chapter 3}

The current Chapter 3 analysis is fully supported by the completed sentence-level annotations. However, if future versions of the report are expected to include exact word accuracy, phoneme-level error distributions, MOS-style ratings, listener-comfort metrics, or task-performance evidence, the following additional questions should be added to the annotation protocol.

\subsection*{A. Word-Level Accuracy}

\begin{enumerate}
\item For each word in the sentence, mark one label: \textit{Correct}, \textit{Acceptable minor deviation}, or \textit{Incorrect}.
\item If a word is marked incorrect, specify whether it was \textit{omitted}, \textit{substituted}, \textit{mispronounced}, or \textit{inserted}.
\item Did the system misread any suffix, postposition, inflectional marker, or function word?
\end{enumerate}

\subsection*{B. Phoneme-Level Error Tagging}

\begin{enumerate}
\item If there is a pronunciation error, which category best describes it: \textit{vowel distortion}, \textit{aspiration loss}, \textit{retroflex/dental confusion}, \textit{voicing confusion}, \textit{conjunct simplification}, \textit{schwa insertion/deletion}, \textit{word-final consonant drop}, \textit{nasalization error}, or \textit{stress misplacement}?
\item Which exact word or phoneme sequence triggered the error?
\item Did the error change meaning, or was it only perceptually unnatural?
\end{enumerate}

\subsection*{C. Numeric and Symbol Normalization}

\begin{enumerate}
\item Was each numeral rendered correctly as spoken Bangla? Mark \textit{Correct}, \textit{Partially Correct}, or \textit{Incorrect}.
\item If incorrect, which numeric type failed: \textit{year}, \textit{date}, \textit{phone number}, \textit{percentage}, \textit{currency}, \textit{decimal}, or \textit{ordinal}?
\item Were symbols such as \texttt{+}, \texttt{-}, \texttt{\#}, \texttt{@}, and \texttt{\%} verbalized correctly?
\end{enumerate}

\subsection*{D. MOS-Style Perceptual Scoring}

\begin{enumerate}
\item Rate naturalness on a 1--5 scale, where 1 is \textit{very robotic} and 5 is \textit{human-like}.
\item Rate prosodic appropriateness on a 1--5 scale, considering intonation, pause placement, clause grouping, and sentence-final contour.
\item Rate fluency on a 1--5 scale, considering pacing, micro-pauses, audio glitches, and synthesis smoothness.
\end{enumerate}

\subsection*{E. User Satisfaction}

\begin{enumerate}
\item How comfortable would you feel listening to this voice for one minute of continuous content? Rate 1--5.
\item Would you trust this voice for educational or assistive reading? Mark \textit{Yes}, \textit{Maybe}, or \textit{No}.
\item If both voices are available, which one would you prefer for regular use: \textit{Male}, \textit{Female}, or \textit{Equal}?
\end{enumerate}

\subsection*{F. Task Performance}

\begin{enumerate}
\item After hearing the sample once, can the listener repeat the key number, date, or name correctly?
\item After hearing the sample once, can the listener answer one factual comprehension question about the sentence?
\item If the sample contains an instruction, can the listener restate the intended action correctly?
\end{enumerate}
